MLH1 Results Outline

### 2020-02-21 doc/Results.rmd
wacky idea about nrm IFDs

---
title: "MLH1 Results Outline"
date: 
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '5'
    toc_float: yes
  html_notebook: default
  pdf_document: default
  word_document:
    toc: yes
    toc_depth: '5'
bibliography: refs_Main_Report.bib
---

@@

< note about the rates of 0CO, 3CO and 4CO? these are marginal chromsomes classes / we'll discuss later?) >

@@

<SUMMARY / TRANSITION to single bivalen measures>

@@

From our total set of cell images `r length(new.Curate.FULL.org$Obj.ID)` chromosome objects were isolated by the image analysis software. After the human curation step, `r length(new.Curate.FULL.org$Obj.ID)` single bivalent measures were left. (make sure these are the hand.foci bivalents).

how do these agree with other patterns of MLH1

@@

#sort by category,

#compare total boxes and rate passed
#Curated_BivData  -- pass rate?

#number of mice, number of cells
#mean number of isolate bivs per cell,

@@

Table X illustrates the break down of single bivalent observations by category 
Error compared to human measures and other details have been discribed in (Peterson_2019).

While the automated software doesn't isolate all bivalents/chromosomes from each cell (on average 17), we assume that the isolation process is not biased. Because there are hundreds of observations per category, we assume that each of the 19 autosomes (chromosomes) is equally represented in the dataset of single bivalents.

@@

These questions use different datasets: Q1 uses a dataset with only sex matched strains  and Q2 uses the data set with all male observations with including strains including those not in the Q1 set.

The same basic models from the MLH1 counts per cell will be used. In addition to basic t.tests and logistic regression models for Q2 to distinguish betwee high and low recombining strains.

# Q1 SC Lengths

@@

EXPLAIN WHY USING TOTAL SC AREA --

turn out comparing the means of single bivalents is complicated by the fact we can't identify any of the bivalents. -- so we use total sc

@@

Q1.real.short_mouse_table <- ddply(real.short.bivData, c("mouse"), summarise,
                        n_obs = length(Obj.ID),
                         mean_SC = mean(chromosomeLength)
                         )

@@

Q1.real.long_mouse_table <- add_strain(Q1.real.long_mouse_table)
Q1.real.long_mouse_table <- add_sex(Q1.real.long_mouse_table)
Q1.real.long_mouse_table <- add_subsp(Q1.real.long_mouse_table)

The plots below show the distributions of mouse average total SC area.

@@

```{r Q1.tot.SC_show, echo=FALSE}

@@

@@

Tot.SC_mouse.av_Musc
```

these are the number of mice for each of the categories

`r as.data.frame( table(droplevels( totSC.mouse.av.table$category) ) )`

@@

The full data set of SC lengths illustrates the general pattern than females have longer SC-AE.

The most averages for total SC (skeletonizes) are significantly longer in females for all strains (p=
`r format(Q1_tot.sc_t.test.WSB$p.value, scientific = F )`, `r format(Q1_tot.sc_t.test.G$p.value, scientific= F )`, and `r format(Q1_tot.sc_t.test.LEW$p.value, scientific= F )`
for the _domesticues^WSB^ _, _domesticues^G^ _, and _domesticues^LEW^ _ respectively).

@@

$$mouse \ average \ total \ SC  \ length ~=~ subsp * sex + rand(strain) + \varepsilon $$

```{r Q1.lmer_totalSC, echo=FALSE, warning=FALSE}

@@

there are more significant effects for strain and subsp for the mixed model of mouse.av total sc

the 3 fixed effects and one random effect were significant (LTR; subsp p= `r Q1.lmer.totSC_results$sub.sp`, 
sex p = `r Q1.lmer.totSC_results$sex.results`, and interaction p = `r Q1.lmer.totSC_results$interaction.pval`, (random LRT; strain p = `r Q1.lmer.totSC_results$rand.pval`)

These effects were followed up the glms

**M2.Linear Model** $$mouse \ average \ total SC  \ area ~=~ sex * strain + \varepsilon $$

**M3. Linear Model** $$mouse \ average \ total SC  \ area ~=~ sex * strain + \varepsilon $$

```{r Q1.glm_totalSC, echo=FALSE, }
#total-SC

#Q1.lmer_totSC.mouse.av.table

Q1.glm_totSC_M2 <- glm(mean.skel ~ subsp*strain * sex, data=Q1.lmer_totSC.mouse.av.table)
#summary(Q1.glm_totSC_M2)
#subsp effect, many strain, #male

#G * male v.sig and male*LEW

Q1.glm_totSC_M3 <- glm(mean.skel ~ strain * sex, data=Q1.lmer_totSC.mouse.av.table)
#summary(Q1.glm_totSC_M3)
#LOT OF SIGNIFICant strain effects in this model

The results from those models was that there were 
...lots of significant effects.  (G male, )

The summary for the total SC area comparison is...

<!-- transition to short-->

The short bivalent dataset was used -- because there are no confounding factors of the XX -- so makes hapier comparisons across male and females

```{r show.short.biv, echo=FALSE}
Q1.SC.short.mouse_scatter_boxplot <- ggplot( Q1.real.short_mouse_table, aes(y = mean_SC, x = sex, color=strain)) + geom_jitter()+ geom_boxplot(alpha=.2) +scale_color_manual(values=colors_of_strains)+facet_wrap(~subsp, scales="free")+ylim(c(20,115) )+ggtitle("Mouse average SC Length \n short bivalent dataset")

The pattern is generally held for comparisons using the reduced single bivalent data sets (short bivlents are the most comparable set, sinceit is unlikely to include the XX in the female data

with the mouse average for 'short' bivalents

(t.test; p = `r Q1_mus.av_shortSC_t.test.WSB$p.value`, `r format(Q1_mus.av_shortSC_t.test.G$p.value, scientific = F)`, and `r format(Q1_mus.av_shortSC_t.test.LEW$p.value, scientific = F)`
for WSB, G and LEW respectively.

(t.test; p= `r format(Q1_mus.av_shortSC_t.test.PWD$p.value, scientific = F)`, `r Q1_mus.av_shortSC_t.test.KAZ$p.value`, and `r  format(Q1_mus.av_shortSC_t.test.SKIVE$p.value, scientific = F)` for PWD, KAZ, SKIVE

for the short bivalents are not significantly differen, for this strain there are 4 female and 4 male

** there are more SKIVE female mouse observations -- but they are filtered out because I excluded them from the MLH1 dataset.  -- for low quality -- they might have good total SC spreads -- // but

* it think total sc only has 1 female observation but the short bivalent has 4 skive females

Mixed models for the sex differences in mouse averages for short bivalents (reduced dataset)

**M2.** $$mouse \ average \ short \ SC  \ length ~=~ subsp * sex * strain + \varepsilon $$

```{r Q1.SC.Length_M2, echo=FALSE, warning=FALSE}

#skive strin very slight sign
#male highly sig
#skive * male slightly sig
```

skive strin very slight sign, male highly sig,  skive * male slightly sig

**M3.** $$mouse \ average \ short \ SC  \ length ~=~ subsp * sex + rand(strain) + \varepsilon $$

```{r Q1_short_biv_lmer, echo=FALSE, warning=FALSE}
lmer.Q1_SCLength_short_M1 <- lmer(mean_SC ~ subsp * sex + (1|strain), data=Q1.real.short_mouse_table, REML = TRUE)

@@

Sex is the most significant factor for the mixed model of mean short bivalents (LTR; p =`r Q1lmer.MM_SC_short_results$sex.results`). interaction effect also significant (p = `r Q1lmer.MM_SC_short_results$interaction.pval`)

**M3.** $$mouse \ average \ short \ SC  \ length ~=~ sex * strain + \varepsilon $$

```{r Q1.glm_short.SC_M3, echo=FALSE}

@@

summary(Q1.glm_short.SC_M3)

@@

round(summ.Q1.glm_short.SC_M3$coefficients[,c(1,4)],5)
#sex male very sig -- only sig
```

male is the only significant p value from this model.

<!--
Because of confonding effects across chromsomes of different physical lenghts, we attempt to ... reduce the error from mixing chromosomes by constructing reduced bivalent sets, for the longest 5 and shortest 5 bivalents from the same cell. The mouse means for these two data sets were then calculated.
-->

Above are the p values for the components of the mixed models for the short and long biv mouse means. In both sets of models the sex effect is the most significant p value.

Above are the mixed model results seperated by mouse averages for 1 crossover and 2 crossover bivalents respectively.

The fixed sex effect is highly significant for the mouse means of SC length. The data sets tested were all pooled SC lengths, the long bivalents, 1CO and 2CO seperately.

In summary .... females (oocytes) have different chromatin compaction patterns compared to males.

@@

# Q1. General CO Positions

@@

We focus on the foci positions from one crossover (1CO) bivalents since the landscape patterns for multi crossover bivalents will be highly influences by crossover interference. We use the normalized foci positions (smaller meaning closer to the centromere and larger being cloerse to the telomere ). (all chromosomes in house mouse are acrocentric)

Chromosome size effects are a confonding factor for CO position, with shorter chromosomes having a more uniform landscape compared to larger chromosomes (cite). To account for this we again use the reduct bivalent sets (long.biv and short.biv)

Below are the general patterns / plots -- short and long bivalent groups are shown to try and account for the chromsomes size effect.

@@

```{r Q1show.F1pos_Full, echo=FALSE}

@@

In the domesticus strains the mouse averages for normalized single foci positions were significantly sexually dimorphic (with female)
are highly significantly more medially placed (t.test; p = `r Q1_1CO.pos_t.test.WSB$p.value`, `r Q1_1CO.pos_t.test.G$p.value`, and `r Q1_1CO.pos_t.test.LEW$p.value` for WSB, G and LEW respectively). With the musclus and molossinus strains displaying a similar pattern (t.test p = `r Q1_1CO.pos_t.test.PWD$p.value` and `r Q1_1CO.pos_t.test.MSM$p.value`) PWD and MSM. Two musculus  strains had a weaker signal, but it was still in the same direction (t.test; p =  `r Q1_1CO.pos_t.test.KAZ$p.value`, and `r Q1_1CO.pos_t.test.SKIVE$p.value` for KAZ and SKIVE respectively).

<!-- lmer and glm here -->

@@

```{r Q1.CO1pos_lmer.M1, echo=FALSE}

@@

lmer.Q1.nrm1CO_M1_results

@@

```
For the Mixed model, sex and strain are significant effects for the normalized single foci position.

@@

Q1.glm_F1pos_M2 <- glm(norm.first.foci.pos ~ subsp * strain * sex, data=Q1.Foci.pos_DF)
summary(Q1.glm_F1pos_M2)

summry.Q1.glm_F1pos_M2 <- summary(Q1.glm_F1pos_M2)
round(summry.Q1.glm_F1pos_M2$coefficients[,c(1,4)],5)

@@

@@

```{r Q1.CO1pos_glm.M3, echo=FALSE, eval=FALSE}

summry.Q1.glm_F1pos_M2

Q1.glm_1CO.pos_M3 <- glm(norm.first.foci.pos ~ strain * sex, data=Q1.Foci.pos_DF)
summary(Q1.glm_1CO.pos_M3)

summry.Q1.glm_1CO.pos_M3 <- summary(Q1.glm_1CO.pos_M3)
round(summry.Q1.glm_1CO.pos_M3$coefficients[,c(1,4)],5)

#write.table(round(summry.Q1.glm_nrmF1pos_M3$coefficients,5), "~./MLH1repo/summry.Q1.glm_nrmF1pos_M3.csv", sep=",",row.names = TRUE)
```

Lew and KAZ had slightly significant strain speicific effects across the two glms (these results will be explored more fully in the Q2 section). The general pattern that emerges across these models is the sex effect being the most significant, (with the male single foci position being more telomeric compared to females).

<!--
-full 1CO tests

long 1CO tests

short 1CO tests

There are a few traits that fall within the CO positions  
- 1CO normalized position  
- sis.co.ten (sorta interference)  
- centromere and telomere distance  
**Heterochiasmy**

2.A) Normalized 1CO positions will be sexually dimorphic (sedell and Kirkpatrick). 
  B) **sis**ter **co**hesion **ten**sion (sis-co-ten) will be sexually dimorphic as it reflects the general property of unifrom vs telomere/biased CO positioning.
  C) Centromere and telomere distances will be sexually dimorphic.
**for all the test blocks below, integrate the 3 other metrics; siscoten, telomere, centromere distance.
-->

```{r Q1.F1.pos.LONG.SHORT.BIV_scatter, echo=FALSE}

Q1.short.biv_scatr_box   <- ggplot(Q1.short.biv_1CO, aes(x=sex, y=PER_Foci_1, color=strain) )+geom_jitter( )+ geom_boxplot(aes(y= PER_Foci_1 ),alpha=.3 ) + scale_color_manual(values=colors_of_strains)+facet_wrap(~subsp, scales = "free")+ylim(c(0,1.1))+ggtitle("Mouse av. \nNormalized 1CO Foci pos \nShort Biv")
Q1.short.biv_scatr_box

Q1.long.biv_scatr_box   <- ggplot(Q1.long.biv_1CO, aes(x=sex, y=PER_Foci_1, color=strain) )+geom_jitter( )+ geom_boxplot(aes(y= PER_Foci_1 ),alpha=.3 ) + scale_color_manual(values=colors_of_strains)+facet_wrap(~subsp, scales = "free")+ylim(c(0,1.1))+ggtitle("Mouse av. \nNormalized 1CO Foci pos \nLong Biv")
Q1.long.biv_scatr_box
```

@@

#mouse.av.long.biv.table
Q1.av.long.biv_scatr_box   <- ggplot(Q1.long.1COpos_mouse.av, aes(x=sex, y=mean_F1pos_PER, color=strain) )+geom_jitter( )+ geom_boxplot(aes(y= mean_F1pos_PER ),alpha=.3 ) + scale_color_manual(values=colors_of_strains)+facet_wrap(~subsp, scales = "free")+ylim(c(0,1.1))+ggtitle("Mouse av. \nNormalized 1CO Foci pos \nLong Biv")
Q1.av.long.biv_scatr_box

Q1.av.short.biv_scatr_box   <- ggplot(Q1.short.1COpos_mouse.av, aes(x=sex, y=mean_F1pos_PER, color=strain) )+geom_jitter( )+ geom_boxplot(aes(y= mean_F1pos_PER ),alpha=.3 ) + scale_color_manual(values=colors_of_strains)+facet_wrap(~subsp, scales = "free")+ylim(c(0,1.1))+ggtitle("Mouse av. \nNormalized 1CO Foci pos \nShort Biv")
Q1.av.short.biv_scatr_box

Above plot focuses on the 1CO bivalent normalized positions since CO interference controls the general position of COs when there are multiple COs. This plot shows the sexual dimorphism in the density plots.

Consider adding annotate_text for the number of observations in each category.
think about adding a vertical line for centromere, for the position means. Think about removing the extra Musc strains.

<!-- Tired runnin plots for the long biv data set, I think there are too few observations. -->

@@

the distribution of SC lengths and sis-coten seems very different across sexes

The mixed model data should only come from 1CO bivalent data.

the mouse average foci1 pos is more significant in t.test, but not log regression... (is something wrong?) Check the mouse averages for the F1_pos, there might be an outlier or mouse with v.few observations.

<!--
### Siscoten

The metric Sis-co-ten measures the amount of sister cohesion connected to the other pole.

>>>>

---
title: "MLH1 Results Outline"
date: 
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '5'
    toc_float: yes
  pdf_document: default
  word_document:
    toc: yes
    toc_depth: '5'
bibliography: refs_Main_Report.bib
---

@@

<note about the rates of 0CO, 3CO and 4CO? these are marginal chromsomes classes / we'll discuss later?)>

@@

Breif conclusion: the evolution of genome wide recombiantion rates proceeds through the chromosome level.

<something about the female results>

<SUMMARY the / TRANSITION to single bivalen measures>

@@

From our total set of cell images `r length(new.Curate.FULL.org$Obj.ID)` chromosome objects were isolated by the image analysis software. After the human curation step, `r length(Curated_BivData$Obj.ID)` single bivalent measures were left. (make sure these are the hand.foci bivalents).

<how do these agree with other patterns of MLH1>

@@

#sort by category, 
#compare total boxes and rate passed
#Curated_BivData  -- pass rate?

#number of mice, number of cells
#mean number of isolate bivs per cell,
#use these two metrics
#cell.level.BivData_table$boxs.IDd 
#cell.level.BivData_table$nboxes_passed

#calq the mean rate by category

cell.level.BivData_table$pass.rate <- cell.level.BivData_table$nboxes_passed / cell.level.BivData_table$boxs.IDd

algorithm.stat.mouse <- ddply(cell.level.BivData_table, c("mouse"), summarise,
                        n_obs = length( unique(fileName) ),
                         mean_pass.rate = mean(pass.rate)
                         )

cell.level.BivData_table <- add_sex(cell.level.BivData_table)

algorithm.stat.category <- ddply(cell.level.BivData_table, c("sex", "strain"), summarise,
                        n_obs = length(unique(fileName)),
                        mean_pass.rate = mean(pass.rate)
                         )

@@

Table X illustrates the break down of single bivalent observations by category. 
Error compared to human measures and other details have been discribed in (Peterson_2019).

The ranges of of rates of isolating chromosomes per cell are `r min(algorithm.stat.category$mean_pass.rate)` in MSM male to 
`r max(algorithm.stat.category$mean_pass.rate)` ) KAZ female, we assume that the isolation process is not biased. Because there are hundreds of observations per category, we assume that each of the 19 autosomes (chromosomes) is equally represented in the dataset of single bivalents. We assume the bivalent datasets across all categories are equally eqivilent for representing general patterns.

<curation, x from curation, XY excluded, replication to hand measures>

@@

<These questions use different datasets: Q1 uses a dataset with only sex matched strains  and Q2 uses the data set with all male observations with including strains including those not in the Q1 set.>
<The same basic models from the MLH1 counts per cell will be used. In addition to basic t.tests and logistic regression models for Q2 to distinguish betwee high and low recombining strains.>

# Q1 Sex Differences in SC-AE Lengths

@@

Comparing the means of single bivalents is complicated by the fact we can't identify the chromosome identitied the physical size (in terms of SC-AE) is the closet approximation we can make.

@@

Q1.real.short_mouse_table <- ddply(real.short.bivData, c("mouse"), summarise,
                        n_obs = length(Obj.ID),
                        ncells = length(unique(fileName)),
                         mean_SC = mean(chromosomeLength)
                         )

@@

Q1.real.long_mouse_table <- add_strain(Q1.real.long_mouse_table)
Q1.real.long_mouse_table <- add_sex(Q1.real.long_mouse_table)
Q1.real.long_mouse_table <- add_subsp(Q1.real.long_mouse_table)
```

For this reason we use a reduced bivalent data set made up of chromsomes (below the 2nd quartile for SC length of single cells). A total of `r sum(Q1.real.short_mouse_table$n_obs)` 'short' bivalents were isolated from `r sum(Q1.real.short_mouse_table$ncells)` oocytes and spermatocytes.

Even though this smaller data set has decreased power due to redcued sample size, ...

This 'short biv' data set has advantages in that it is closer to sampling across a similar set of chromosomes (ideally Chromosomes 15 to 19) and it excludes the XX in females, which is predicted to be the 3rd longest based on the physical (Mb) length.

@@

```{r Q1.SC.show.short.biv, echo=FALSE}
Q1.SC.short.mouse_scatter_boxplot <- ggplot( Q1.real.short_mouse_table, aes(y = mean_SC, x = sex, color=strain)) + geom_jitter()+ geom_boxplot(alpha=.2) +scale_color_manual(values=colors_of_strains)+facet_wrap(~subsp, scales="free")+ylim(c(20,115) )+ggtitle("Mouse average SC Length \n short bivalent dataset")

<in the above boxplots of mouse means for shortbivalents>

For all but one strain, _musculus^SKIVE^_, the mouse mean for the short bivalents are significantly longer <how much longer, the largest difference across sexes by strain are>  (t.test; p = `r Q1_mus.av_shortSC_t.test.WSB$p.value`, `r format(Q1_mus.av_shortSC_t.test.G$p.value, scientific = F)`, and `r format(Q1_mus.av_shortSC_t.test.LEW$p.value, scientific = F)`
for _domesticus^WSB^_, _domesticus^G^_ and _domesticus^LEW^_ respectively.
and (t.test; p= `r format(Q1_mus.av_shortSC_t.test.PWD$p.value, scientific = F)`, `r Q1_mus.av_shortSC_t.test.KAZ$p.value`, and `r  format(Q1_mus.av_shortSC_t.test.SKIVE$p.value, scientific = F)` for _musculus^PWD^_ , _musculus^KAZ^_ , _musculus^SKIVE^_  respectively.

<EXPLAIN WHY USING TOTAL SC AREA>

An additional measure of the SC-AE we tested is the total SC area per cell (the whole genome).**to get a comparison of the SC-AE area for the whole genome (not just a subset of chromosomes)** We apply a skeletonizing image analysis algorithm to all MLH1 meiocyte spreads. A total of X images were run trhought the image analysis pipeline, and after outliers removed, X remained. / cell based measures were left. Mouse means were calculated for their total SC-AE area and are displayed in figure X with the strain means shown as horizontal lines (below).

```{r Q1.tot.SC_show, echo=FALSE, warning=FALSE}

@@

  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank())+  ggtitle("Mouse av. total SC area")+  ylim(c(1000,3000 ))

Tot.SC_mouse.av_Musc

totsc.mol <- Q1.totSC.mouse.av.table %>% filter( subsp == "Mol" )
totsc_strain.av.mol <- totSC.strain.av.table %>% filter( subsp == "Mol" )

Tot.SC_mouse.av_Mol <- ggplot(data= totsc.mol, aes(y = mean.skel, x=mouse, color=strain))+ geom_point()+scale_color_manual(values=colors_of_strains)+
  facet_wrap(~sex, scales="free", ncol = 3)+

  geom_hline(data =totsc_strain.av.mol, aes(yintercept=mean.skel) )+

@@

Tot.SC_mouse.av_Mol
```

<these are the number of mice for each of the categories r as.data.frame( table(droplevels( totSC.mouse.av.table$category) ) )

@@

```{r Q1.totSC_male.first.comp, echo=FALSE}

Q1.male.Dom.Musc.t.test <- t.test( ( totSC.mouse.av.table %>% filter(sex == "male") %>% filter(subsp == "Dom") )$mean.skel,
(totSC.mouse.av.table %>% filter(sex == "male") %>% filter(subsp == "Musc") )$mean.skel)

Q1.male.Dom.Mol.t.test <- t.test( ( totSC.mouse.av.table %>% filter(sex == "male") %>% filter(subsp == "Dom") )$mean.skel,
(totSC.mouse.av.table %>% filter(sex == "male") %>% filter(subsp == "Mol") )$mean.skel)
```

The most averages for total SC-AE area (skeletonizes) are significantly longer in females for all strains (p=
`r format(Q1_tot.sc_t.test.WSB$p.value, scientific = F )`, `r format(Q1_tot.sc_t.test.G$p.value, scientific= F )`, and `r format(Q1_tot.sc_t.test.LEW$p.value, scientific= F )`
for the _domesticues^WSB^ _, _domesticues^G^ _, and _domesticues^LEW^ _ respectively).

@@

The comparison of this metric is somewhat complicated across sexes due to the sex chromosomes. (simply put, the oocytes have an additional autosome, and the spermatocytes have mroe mostly non-recombinationly active SC-AE area, the XY).  <But we use this metric anyways>

<We estimate the effect of the sex chromosomes from hand measured data...the XX (3rd longest chromsomes) in females is ~10% of total the summed SC area, the average propotion of the XY in spermatocytes is XX%?  however the morphology of the XY is highly variable across cells>

The mouse means for total SC in domesticus strains are significant shorter compared to Musc (t.test; p= `r Q1.male.Dom.Musc.t.test$p.value`) and Molossinus (t.test; p = `r Q1.male.Dom.Mol.t.test$p.value`, indecating evolution (divergence) of SC length and (chromatin compaction) <this will be expanded on n later sections>
-- we follow up with models (which focus on sex differences).

<explain why applying models.  We then use models to examine/ think about the evolution>

The above t.tests were done to establish the sex differences for SC-AE (chromatin condensation), we now apply models (for understanding within an evolutionary framework) -- we apply the same models to both mean mouse total sc and mouse mean short bivalent.

**M3. Mixed Model** $$mouse \ average \ SC \ metric ~=~ subsp * sex + rand(strain) + \varepsilon $$

```{r Q1.lmer_totalSC, echo=FALSE, warning=FALSE, message=FALSE}

@@

For the mixed model run on the total sc metric, there are more significant effects for strain and subsp for the mixed model of mouse.av total sc.

The three fixed effects and one random effect were significant (LTR; subsp p= `r Q1.lmer.totSC_results$sub.sp`, 
sex p = `r Q1.lmer.totSC_results$sex.results`, and interaction p = `r Q1.lmer.totSC_results$interaction.pval`, (random LRT; strain p = `r Q1.lmer.totSC_results$rand.pval`).

```{r Q1_short_biv_lmer, echo=FALSE, warning=FALSE, message=FALSE}
lmer.Q1_SCLength_short_M1 <- lmer(mean_SC ~ subsp * sex + (1|strain), data=Q1.real.short_mouse_table, REML = TRUE)

@@

<and the short biv>
Mixed models for the sex differences in mouse averages for short bivalents Sex is the most significant factor for the mixed model of mean short bivalents (LTR; p =`r Q1lmer.MM_SC_short_results$sex.results`). interaction (subspecies by sex) effect also significant (p = `r Q1lmer.MM_SC_short_results$interaction.pval`).

These effects were followed up the glms, again applied to both mouse means for short biv SC lengths and total sc area per cell.

**M2.Linear Model** $$mouse \ average \ SC \ metric ~=~ sex * strain + \varepsilon $$

**M3. Linear Model** $$mouse \ average \ SC \ metric ~=~ sex * strain + \varepsilon $$

```{r Q1.glm_totalSC, echo=FALSE, message=FALSE}
#total-SC
#Q1.lmer_totSC.mouse.av.table

Q1.glm_totSC_M2 <- glm(mean.skel ~ subsp*strain * sex, data=Q1.lmer_totSC.mouse.av.table)
sum.Q1.glm_totSC_M2 <- summary(Q1.glm_totSC_M2)

#subsp effect, many strain, #male
#G * male v.sig and male*LEW

Q1.glm_totSC_M3 <- glm(mean.skel ~ strain * sex, data=Q1.lmer_totSC.mouse.av.table)
sum.Q1.glm_totSC_M3 <- summary(Q1.glm_totSC_M3)
#LOT OF SIGNIFICant strain effects in this model
```

The following effects had significant p values given the above model

subspecies musc p = `r sum.Q1.glm_totSC_M2$coefficients[,4][2]`, subspecies mol p =`r sum.Q1.glm_totSC_M2$coefficients[,4][3]`, sex male  p =`r sum.Q1.glm_totSC_M2$coefficients[,4][8]`, strain lew p =`r sum.Q1.glm_totSC_M2$coefficients[,4][5]`, strain g  p =`r sum.Q1.glm_totSC_M2$coefficients[,4][4]`,
<this is the smallest p vale, elevated female total, driven by the high means in females>

(subsp*sex) mol.male  p =`r sum.Q1.glm_totSC_M2$coefficients[,4][10]`, sex.g  p =`r sum.Q1.glm_totSC_M2$coefficients[,4][11]`, lew.male p =`r sum.Q1.glm_totSC_M2$coefficients[,4][12]`

These results indicate that there are strain specific effects for the SC-AE area --

the large G strain effect is likely driven by the high female mean total SC area

sex is the next strongest (most significant) effect.

For the short bivalent data -- run through the same models...

```{r Q1.SC.Length_M2, echo=FALSE, warning=FALSE, message=FALSE}

```{r Q1.glm_short.SC_M3, echo=FALSE, message=FALSE}

@@

#summary(Q1.glm_short.SC_M3)

@@

#round(summ.Q1.glm_short.SC_M3$coefficients[,c(1,4)],5)
#sex male very sig -- only sig
```

From this (M2) model sex effect, male is most significant (glm; p =`r summ.Q1.glm_short.SC_M2$coefficients[,4][10]`). while the skive strain effect (glm; p=`r summ.Q1.glm_short.SC_M2$coefficients[,4][8]`) and and SKIVE*male (glm; p =`r summ.Q1.glm_short.SC_M2$coefficients[,4][16]`) are slightly significant.

In the M2 model, the Sex effect (male) is the only significant p value from this model (glm; p= `r summ.Q1.glm_short.SC_M3$coefficients[,4][10]`).
<should apply caution to the short bivalent dataset> The significant SKIVE effect is driven by  2 of the 4 female skive mice -- having much shorter bivalents comapred to the other female musc strain means.

<SUMMARY, ttests and models indicate female have strongest effect -- longer AC-AE.  >

In summary, females (oocytes) have loser different chromatin compaction patterns compared to males. There are suggestions of evolution for this trait across the domesticus and musculus subspecies.

@@

# short transition

<transition>  -- following up on SC area - with how COs are place --- placement and number of COs on single bivalents
1. Transition, --- the SC-AE area can be thought of as the 'area / available real-estate for COs
-- these results suggest that this area is 'utilized' in different ways across sexes and strains - so we next investigate the recombination landscape - (ie the relationship between the placement and number of COs along chromosomes).

2. We focus on two metrics/ aspects i) the placement of single foci along a bivalent (just 1CO) and ii) the placement of two foci on the same bivalent / the interfocal distance of 2CO bivalents (reflects crossover interference)

# Q1. 1CO Position Sex Differences

@@

We focus on the foci positions from one crossover (1CO n= 
) bivalents since the landscape patterns for multi crossover bivalents will be highly influences by crossover interference.

We anchor the position measures at the centromere (and normalize by deviding the position by the length of the chromosome. The foci positions then have a range of 0-1, with smaller values representing positions closer to the centromere end of the bivalent, and larger values meaning foci positions closer to the opposite end (refered here as telomeric position).

We plot the mouse averages across all of the single bivalent measures, with the strain averages shown as boxplots...

in order to get a better picture of the sex differences in this foci-position metric

@@

```{r Q1show.F1pos_Full, echo=FALSE, warning=FALSE}

@@

In the domesticus strains the mouse averages for normalized single foci positions were significantly sexually dimorphic (with female)
are highly significantly more medially placed (t.test; p = `r Q1_1CO.pos_t.test.WSB$p.value`, `r Q1_1CO.pos_t.test.G$p.value`, and `r Q1_1CO.pos_t.test.LEW$p.value` for WSB, G and LEW respectively). With the musculus and molossinus strains displaying a similar pattern (t.test p = `r Q1_1CO.pos_t.test.PWD$p.value` and `r Q1_1CO.pos_t.test.MSM$p.value`) PWD and MSM. Two musculus  strains had a weaker signal, but it was still in the same direction (t.test; p =  `r Q1_1CO.pos_t.test.KAZ$p.value`, and `r Q1_1CO.pos_t.test.SKIVE$p.value` for KAZ and SKIVE respectively).

After confirming the significan sex differences for 1CO position, we apply the familiar models

**M1. Mixed Model**

@@

```{r Q1.CO1pos_lmer.M1, echo=FALSE, message=FALSE}

@@

#lmer.Q1.nrm1CO_M1_results

@@

For the Mixed model, sex is the most significant effect (LTR; p =`r lmer.Q1.nrm1CO_M1_results$sex.results`).The random effect of strain is also significant
(LRT; p =`r lmer.Q1.nrm1CO_M1_results$rand.pval`). These results confirm the sex difference establish in the t.tests above -- and suggesting variation across the strains (in the sexual dimorphism for this pattern).

**M2. Linear model**

@@

Q1.glm_F1pos_M2 <- glm(norm.first.foci.pos ~ subsp * strain * sex, data=Q1.Foci.pos_DF)

#summary(Q1.glm_F1pos_M2)

summry.Q1.glm_F1pos_M2 <- summary(Q1.glm_F1pos_M2)
#round(summry.Q1.glm_F1pos_M2$coefficients[,c(1,4)],5)

@@

**M3. Linear model**

@@

```{r Q1.CO1pos_glm.M3, echo=FALSE, eval=TRUE}

Q1.glm_1CO.pos_M3 <- glm(norm.first.foci.pos ~ strain * sex, data=Q1.Foci.pos_DF)

#summary(Q1.glm_1CO.pos_M3)

summry.Q1.glm_1CO.pos_M3 <- summary(Q1.glm_1CO.pos_M3)
#round(summry.Q1.glm_1CO.pos_M3$coefficients[,c(1,4)],5)
#write.table(round(summry.Q1.glm_nrmF1pos_M3$coefficients,5), "~./MLH1repo/summry.Q1.glm_nrmF1pos_M3.csv", sep=",",row.names = TRUE)

The male effects were the most significant (glm M2; p =`r summry.Q1.glm_F1pos_M2$coefficients[,4][8]`, glm M3; p = `r summry.Q1.glm_1CO.pos_M3$coefficients[,4][8]`). While the musc subspecies, LEW and KAZ strains had slightly signifcant effects (glm M2; Musc subsp p = `r summry.Q1.glm_F1pos_M2$coefficients[,4][2]`, LEW strain p = `r summry.Q1.glm_F1pos_M2$coefficients[,4][5]`) and (glm M3; LEW strain p = `r summry.Q1.glm_1CO.pos_M3$coefficients[,4][3]`, KAZ strain p = `r summry.Q1.glm_1CO.pos_M3$coefficients[,4][7]`).

(these results will be explored more fully in the Q2 section). The general pattern that emerges across these models is the sex effect being the most significant, (with the male single foci position being more telomeric compared to females).

<we don't explore the effects on the redcued bivalent data sets! The main expected pattern is confirmed>

Chromosome size effects are a confounding factor for CO position, with the expectation that shorter chromosomes will have a more uniform landscape compared to larger/longer chromosomes (cite). To account for this we again use the reduced bivalent sets (long.biv and short.biv)

```{r Q1.F1.pos.LONG.SHORT.BIV_scatter, echo=FALSE}
Q1.short.biv_scatr_box   <- ggplot(Q1.short.biv_1CO, aes(x=sex, y=PER_Foci_1, color=strain) )+geom_jitter( )+ geom_boxplot(aes(y= PER_Foci_1 ),alpha=.3 ) + scale_color_manual(values=colors_of_strains)+facet_wrap(~subsp, scales = "free")+ylim(c(0,1.1))+ggtitle("Mouse av. \nNormalized 1CO Foci pos \nShort Biv")

Q1.long.biv_scatr_box   <- ggplot(Q1.long.biv_1CO, aes(x=sex, y=PER_Foci_1, color=strain) )+geom_jitter( )+ geom_boxplot(aes(y= PER_Foci_1 ),alpha=.3 ) + scale_color_manual(values=colors_of_strains)+facet_wrap(~subsp, scales = "free")+ylim(c(0,1.1))+ggtitle("Mouse av. \nNormalized 1CO Foci pos \nLong Biv")

@@

#mouse.av.long.biv.table
Q1.av.long.biv_scatr_box   <- ggplot(Q1.long.1COpos_mouse.av, aes(x=sex, y=mean_F1pos_PER, color=strain) )+geom_jitter( )+ geom_boxplot(aes(y= mean_F1pos_PER ),alpha=.3 ) + scale_color_manual(values=colors_of_strains)+facet_wrap(~subsp, scales = "free")+ylim(c(0,1.1))+ggtitle("Mouse av. \nNormalized 1CO Foci pos \nLong Biv")

Q1.av.short.biv_scatr_box   <- ggplot(Q1.short.1COpos_mouse.av, aes(x=sex, y=mean_F1pos_PER, color=strain) )+geom_jitter( )+ geom_boxplot(aes(y= mean_F1pos_PER ),alpha=.3 ) + scale_color_manual(values=colors_of_strains)+facet_wrap(~subsp, scales = "free")+ylim(c(0,1.1))+ggtitle("Mouse av. \nNormalized 1CO Foci pos \nShort Biv")

@@

<!--
### Siscoten
The metric Sis-co-ten measures the amount of sister cohesion connected to the other pole.
The logic of how the sis-co-ten metric is outlined in the figure below. The goal is to use this metric to model different tension active cohesion amounts as a consequence of different numbers and placements of chiasmata/CO. This metric is calculated using SC area as a proxy to the amount of cohesion at metaphase.
